%%% TO DO %%%
% implement siunitx package and change units

\documentclass[12pt]{article}

\usepackage{ishn}
\usetikzlibrary{decorations.markings}
\usetikzlibrary{patterns}
\usetikzlibrary{patterns.meta}
\usepackage{caption}
\usepackage{subcaption}

\makeindex[intoc]

\begin{document}
\robustify\dots
\sisetup{input-digits = 0123456789\dots}

\tikzset{middlearrow/.style={
        decoration={markings,
            mark= at position 0.5 with {\arrow{#1}} ,
        },
        postaction={decorate}
    }
}

\hypersetup{pageanchor=false}
\begin{titlepage}
	\begin{center}
		\vspace*{1em}
		\Huge
		\textbf{IB Electromagnetism}

		\vspace{1em}
		\large
		Ishan Nath, Lent 2023

		\vspace{1.5em}

		\Large

		Based on Lectures by Prof. Gordon Ogilvie

		\vspace{1em}

		\large
		\today
	\end{center}
	
\end{titlepage}
\hypersetup{pageanchor=true}

\tableofcontents

\newpage

\section{Introduction}
\label{sec:introduction}

\subsection{Charges and Currents}
\label{sub:charges_and_currents}

\emph{Electric charge}\index{electric charge} is a physical property of elementary particles. It is:
\begin{itemize}
	\item Positive, negative or zero.
	\item Quantized (an integer multiple of the \emph{elementary charge} $e$).
	\item Conserved (even if particles are created or destroyed).
\end{itemize}

By convention, the electron has charge $-e$, the proton has charge $+e$, and the neutron has charge $0$.

On macroscopic scales, the number of particles is so large that charge can be considered to have continuous \emph{electric charge density}\index{electric charge density} $\rho(\mathbf{x},t)$. The total charge in a volume $V$ is then
\[
Q = \int_{V} \rho \diff V
.\]

The \emph{electric current density}\index{electric charge density} $\mathbf{J}(\mathbf{x}, t)$ is the flux of electric charge per unit area. The current flowing through a surface $S$ is
\[
I = \int_{S} \mathbf{J} \cdot \diff \mathbf{S}
.\]

Consider a time-independent volume $V$ with boundary $S$. Since charge is conserved, we have
\begin{align*}
	\frac{\diff Q}{\diff t} &= -I, \\
	\frac{\diff}{\diff t} \int_{V} \rho \diff V + \int_{S} \mathbf{J} \cdot d \mathbf{S} &= 0, \\
	\int_{V} \biggl( \frac{\partial \rho}{\partial t} + \nabla \cdot \mathbf{J} \biggr) \diff V &= 0.
\end{align*}
Since this is true for any $V$, we must have
\[
\frac{\partial \rho}{\partial t} + \nabla \cdot \mathbf{J} = 0
.\]

This \emph{equation of charge conservation}\index{equation of charge conservation} has the typical form of a conservation law.

The discrete charge distribution of a single particle of charge $q_i$, and position vector $\mathbf{x}_i(t)$ is
\begin{align*}
	\rho &= q_i \delta( \mathbf{x} - \mathbf{x}_i(t)), \\
	\mathbf{J} &= q_i \mathbf{\dot x}_i \delta(\mathbf{x} - \mathbf{x}_i(t)).
\end{align*}
For $N$ particles, it is
\begin{align*}
	\rho &= \sum_{i = 1}^{N} q_i \delta(\mathbf{x} - \mathbf{x}_i(t)), \\
	\mathbf{J} &= \sum_{i = 1}^{N} q_i \mathbf{\dot x}_i \delta(\mathbf{x} - \mathbf{x}_i(t)).
\end{align*}
We can verify that these distributions satisfy the charge conservation equation.

\subsection{Fields and Forces}
\label{sub:fields_and_forces}

Electromagnetism is a \emph{field theory}\index{field theory}. Charged particles interact not directly, but by generating fields around them that are experienced by other charged particles.

In general, we have two time-dependent vector fields: the \emph{electric field}\index{electric field} $\mathbf{E}(\mathbf{x}, t)$, and the \emph{magnetic field}\index{magnetic field} $\mathbf{B}(\mathbf{x}, t)$.

The \emph{Lorentz force}\index{Lorentz force} on a particle of charge $q$ and velocity $\mathbf{v}$ is
\[
\mathbf{F} = q(\mathbf{E} + \mathbf{v} \times \mathbf{B})
.\]

\subsection{Maxwell's equations}
\label{sub:maxwells_equations}

In this course we will explore some consequences of \emph{Maxwell's equations}\index{Maxwell's equations}
\begin{align*}
	\nabla \cdot \mathbf{E} &= \frac{\rho}{\epsilon_0}, & \nabla \cdot \mathbf{B} &= 0, \\
	\nabla \times \mathbf{E} &= - \frac{\partial \mathbf{B}}{\partial t}, & \nabla \times \mathbf{B} &= \mu_0 \biggl( \mathbf{J} + \epsilon_0 \frac{\partial \mathbf{E}}{\partial t} \biggr).
\end{align*}
Some properties of Maxwell's equations are:
\begin{itemize}
	\item They are coupled linear PDE's in space and time.
	\item They involve two positive constants: $\epsilon_0$ (vacuum permittivity)\index{vacuum permittivity}, and $\mu_0$ (vacuum permeability)\index{vacuum permeability}.
	\item Charges $(\rho)$ and currents $(\mathbf{J})$ are the sources of the electromagnetic fields.
	\item Each equation has an equivalent integral form, related via the divergence theorem of Stokes' theorem.
	\item These are the vacuum equations that apply on microscopic scales (or in a vacuum). A related macroscopic version applies in media (for examples air).
	\item The equations are consistent with each other and with charge conservation. For example, $\nabla \cdot (M3) = \frac{\partial}{\partial t}(M2)$, and
		\[
		\frac{\partial \rho}{\partial t} + \nabla \cdot \mathbf{J} = \frac{\partial}{\partial t} (\epsilon_0 \nabla \cdot \mathbf{E}) + \nabla \cdot \biggl(- \epsilon_0 \frac{\partial \mathbf{E}}{\partial t} + \frac{1}{\mu_0} \nabla \times \mathbf{B} \biggr) = 0
		.\]
\end{itemize}

\subsection{Units}
\label{sub:units}

The SI unit of electric charge is the coulomb (\unit{\coulomb}). The elementary charge is (exactly)
\[
	e = \qty{1.602176634e-19}{\coulomb}
.\]
The SI unit of electric current is the ampere, or amp (\unit{\ampere}), equal to \qty{1}{\coulomb\per\second}.

The SI base units needed in electromagnetism are:
\begin{itemize}
	\item[] second (\unit{\second})
	\item[] metre (\unit{\metre})
	\item[] kilogram (\unit{\kilogram})
	\item[] ampere (\unit{\ampere})
\end{itemize}

From the Lorentz force law, we can see that the units of $\mathbf{E}$ and $\mathbf{B}$ must be
\begin{center}
	\unit{\kilogram\metre\per\second\cubed\per\ampere} and \unit{\kilogram\per\second\squared\per\ampere}.
\end{center}
The latter is also called the tesla (\unit{\tesla}). From Maxwell's equations, we can work out the units of $\epsilon_0$ and $\mu_0$. The experimentally determined values are
\begin{align*}
	\epsilon_0 &= \qty{8.854\dots e-12}{\per\kilogram\per\metre\cubed\second\tothe{4}\ampere\squared} \\
	\mu_0 &= \qty{1.256\dots e-6}{\kilogram\metre\per\second\squared\per\ampere\squared}\\
	      &\approx 4 \pi \times 10^{-7}\, \unit{\kilogram\metre\per\second\squared\per\ampere\squared}.
\end{align*}

The speed of light is (exactly)
\[
	c = \frac{1}{\sqrt{\mu_0 \eps_0}} = \qty{299792458}{\metre\per\second} \approx \qty{3e8}{\metre\per\second}
.\]

\newpage

\section{Electrostatics}
\label{sec:electrostatics}

In a time-independent situation, Maxwell's equations reduce to
\begin{align*}
	\nabla \cdot \mathbf{E} &= \frac{\rho}{\epsilon_0}, & \nabla \times \mathbf{E} &= \mathbf{0}, \\
	\nabla \cdot \mathbf{B} &= 0, & \nabla \times \mathbf{B} &= \mu_0 \mathbf{J}.
\end{align*}
Since $\mathbf{E}$ and $\mathbf{B}$ are decoupled, we can study them separately.

\emph{Electrostatics}\index{electrostatics} is the study of the electric field generated by a stationary charge distribution
\begin{align*}
	\nabla \cdot \mathbf{E} &= \frac{\rho}{\epsilon_0}, \tag{M1}\label{eq:m1}\\
	\nabla \times \mathbf{E} &= \mathbf{0}. \tag{M3'}\label{eq:m3'}
\end{align*}

\subsection{Gauss' Law}
\label{sub:gauss_law}

Consider a closed surface $S$ enclosing a volume $V$. Integrating (\ref{eq:m1}) over $V$ and using the divergence theorem, we obtain \emph{Gauss' law}\index{Gauss' law}
\[
\int_{S} \mathbf{E} \cdot \diff \mathbf{S} = \frac{Q}{\epsilon_0}
,\]
where $Q = \int_{V} \rho \diff V$ is the total charge in $V$.

Gauss' law is the integral version of (\ref{eq:m1}) and is valid generally. This says that the electric flux of a closed surface is proportional to the total charge enclosed.

In special situations, we can use Gauss' law together with symmetry to deduce $\mathbf{E}$ from $\rho$. By choosing the \emph{Gaussian surface}\index{Gaussian surface} $S$ appropriately.

\subsubsection{Spherical Symmetry}
\label{subsub:spherical_symmetry}

Consider a spherically symmetric charge distribution, $\rho(r)$ in spherical polar coordinates, with total charge $Q$ contained within an outer radius $R$.

To have spherical symmetry, the electric field should have the form
\[
	\mathbf{E} = E(r) \mathbf{e}_r
.\]
This will satisfy (\ref{eq:m3'}), as required. To find $E(r)$, we apply Gauss' law to a sphere of radius $r$. If $r > R$, then
\[
\int_{S}\mathbf{E}\cdot \diff \mathbf{S} = E(r) \int_{S} \mathbf{e}_r \cdot \diff \mathbf{S} = E(r) \int_{S} \diff S = E(r) \, 4 \pi r^2 = \frac{Q}{\epsilon_0}
.\]
Thus, outside of the sphere of radius $R$,
\[
\mathbf{E} = \frac{Q}{4 \pi \epsilon_0 r^2}\mathbf{e}_r
.\]
So the external electric field of a spherically symmetric body depends only on the total charge.

The Lorentz force on a particle of charge $q$ in $r > R$ is
\[
\mathbf{F} = q \mathbf{E} = \frac{Qq}{4 \pi \epsilon_0 r^2} \mathbf{e}_r
.\]
This is the \emph{Coulomb force}\index{Coulomb force} between charged particles. The force is repulsive if the charges have the same sign $(Qq > 0)$ and attractive if they have opposite signs $(Qq < 0)$.

If we take the limit $R \to 0$, we obtain the electric field of a \emph{point charge}\index{point charge} $Q$, corresponding to
\[
\rho = Q \delta(\mathbf{x})
.\]

There is a close analogy between the Coulomb force and the gravitational force between massive particles,
\[
	\mathbf{F} = - \frac{GMm}{r^2} \mathbf{e}_r
.\]
Both involve an inverse-square law, and the product of the charges/masses. However,
\begin{itemize}
	\item While gravity is always attractive, electric forces can be repulsive or attractive.
	\item Gravity is very much weaker than the Coulomb force, e.g. for two protons the ratio of the electric to gravitational forces is
		\[
		\frac{e^2}{4 \pi \epsilon_0 G m_p^2} \approx 10^{36}
		.\]
		On the atomic scale, gravity is irrelevant. But positive and negative charges balance so accurately that on the planetary scale, gravity is dominant.
\end{itemize}

\subsubsection{Cylindrical Symmetry}
\label{subsub:cylindrical_symmetry}

Consider a cylindrically symmetric charge distribution $\rho(r)$ in cylindrical polar coordinates, with total charge $\lambda$ per unit length, contained within an outer radius $R$.

To have cylindrical symmetry,
\[
\mathbf{E} = E(r) \mathbf{e}_r
.\]
To find  $E(r)$ we apply Gauss' law to a cylinder of radius $r$ and arbitrary length $L$. Again, we consider $r > R$. Then, since only the curved part of the cylinder contributes to the flux,
\[
\int_{S}\mathbf{E} \cdot \diff \mathbf{S} = E(r) \int_{S}\mathbf{e}_r \cdot \diff \mathbf{S} = E(r) \int_{S} \diff S = E(r) 2 \pi r L = \frac{\lambda L}{\epsilon_0}
.\]
Thus, we get
\[
\mathbf{E} = \frac{\lambda}{2 \pi \epsilon_0 r} \mathbf{e_r}
.\]
In the limit $R \to 0$, we obtain the electric field of a \emph{line charge}\index{line charge} $\lambda$ per unit length, corresponding to
\[
\rho = \lambda \delta(x) \delta(y)
.\]

\subsubsection{Planar Symmetry}
\label{subsub:planar_symmetry}

We consider a planar charge distribution $\rho(z)$ in Cartesian coordinates, with total charge $\sigma$ per unit area, contained within a region $-d < z < d$ of thickness $2d$. We assume reflectional symmetry, so $\rho(z)$ is even.

To have planar symmetry, we need
\[
\mathbf{E} = E(z) \mathbf{e}_z
,\]
which will satisfy (\ref{eq:m3'}). Reflectional symmetry implies $E(-z) = -E(z)$. To find $E(z)$ for $z > 0$, apply Gauss' law to a ``Gaussian pillbox'' of height $2z$ and arbitrary area $A$. If $z > d$, then
\[
\int_{S}\mathbf{E} \cdot \diff \mathbf{S} = E(z)A - E(-z)A = 2E(z)A = \frac{\sigma A}{\epsilon_0}
.\]
Thus,
\[
\mathbf{E} =
\begin{cases}
	\frac{\sigma}{2 \epsilon_0} \mathbf{e}_z & z > d, \\
	-\frac{\sigma}{2 \epsilon_0} \mathbf{e}_z & $z < -d$.
\end{cases}
\]
In the limit $d \to 0$, we obtain the electric field of a \emph{surface charge}\index{surface charge} $\sigma$ per unit area, corresponding to
\[
\rho = \sigma \delta(z)
.\]

\subsubsection{Surface Charge and Discontinuity}
\label{subsub:surface_charge_and_discontinuity}

Let $\mathbf{n}$ be a unit vector normal to the charged surface, pointing from region 1 to region 2. In our example, $\mathbf{n} = \mathbf{e}_z$.

The discontinuity in $\mathbf{E}$ is given by
\[
	[\mathbf{n} \cdot \mathbf{E}] = \frac{\sigma}{\epsilon_0}
,\]
where $\sigma$ is the surface charge density, and
\[
	[X] = X_2 - X_1
\]
denotes a discontinuity. The tangential components are continuous (they are both 0), so
\[
	[\mathbf{n} \times \mathbf{E}] = \mathbf{0}
.\]

These equation apply to any surface charge (even if the surface is curved an non-uniform).

The first comes from applying Gauss' law to an infinitesimal Gaussian pillbox on the surface.

The second comes from considering an infinitesimal circuit that goes through the surface: in the limit, and by taking all orientations of loops, we can use Stokes' theorem to get the required result.

\subsection{The Electrostatic Potential}
\label{sub:the_electrostatic_potential}

For general $\rho(\mathbf{x})$, we cannot determine $\mathbf{E}(\mathbf{x})$ using Gauss' law alone.

Since $\nabla \times \mathbf{E} = \mathbf{0}$, we know that $\mathbf{E}$ can be written in terms of an \emph{electrostatic potential}\index{electrostatic potential} (or electric potential) $\Phi(\mathbf{x})$
\[
\mathbf{E} = - \nabla \Phi
.\]
The \emph{potential difference}\index{potential difference} (or \emph{voltage}\index{voltage}) between two points $\mathbf{x}_1$ and $\mathbf{x}_2$ is
\[
\Phi(\mathbf{x}_2) - \Phi(\mathbf{x}_1) = \int \diff \Phi = - \int_{\mathbf{x}_1}^{\mathbf{x}_2} \mathbf{E}(\mathbf{x}) \cdot \diff \mathbf{x}
,\]
and is path-independent because $\nabla \times \mathbf{E} = \mathbf{0}$.

The electric force on a particle of charge $q$ is
\[
\mathbf{F} = q \mathbf{E} = - q \nabla \Phi
\]
is a conservative force associated with the potential energy
\[
U(\mathbf{x}) = q \Phi(\mathbf{x})
.\]
(\ref{eq:m1}) implies that $\Phi$ satisfies \emph{Poisson's equation}
\[
- \nabla^2 \Phi = \frac{\rho}{\epsilon_0}
.\]
The solution can be written as an integral (over all space, assuming decay at infinity)
\[
\Phi(\mathbf{x}) = \frac{1}{4 \pi \epsilon_0} \int \frac{\rho(\mathbf{x}')}{|\mathbf{x} - \mathbf{x}'|} \Diff3 \mathbf{x}'
.\]
This is the convolution of $\rho(\mathbf{x})$ with the potential of a unit point charge $\frac{1}{4 \pi \epsilon_0 |\mathbf{x}|}$, which is the solution of
\[
- \nabla^2 \Phi = \frac{\delta (\mathbf{x})}{\epsilon_0}
,\]
satisfying $\Phi \to 0$ as $|\mathbf{x}| \to \infty$.

Note that $\mathbf{E}$ is unaffected if we add an arbitrary constant to $\Phi$. We usually choose this constant such that $\Phi \to 0$ as $|\mathbf{x}| \to \infty$. However if $\rho(\mathbf{x})$ does not decay sufficiently rapidly, this may not be possible. For example, a line charge has $E_r \propto \frac{1}{r}$, so $\Phi \propto \log r$, which does not decay.

\subsubsection{Point Charge}
\label{subsub:point_charge}

The potential due to a point charge $q$ at the origin is
\[
\Phi(\mathbf{x}) = \frac{q}{4 \pi \epsilon_0 |\mathbf{x}|} = \frac{q}{4 \pi \epsilon_0 r}
.\]

\subsubsection{Electric Dipole}
\label{subsub:electric_dipole}

This consists of two equal and opposite charge at difference positions. Without loss of generality, consider charges $-q$ at $\mathbf{x} = \mathbf{0}$ and $+q$ and $\mathbf{x} = \mathbf{d}$.

\begin{figure}[h]
	\centering
	\caption{Electric Dipole}
	\label{fig:electric_dipole}
	\vspace{1em}
	\begin{tikzpicture}
		\draw [|-|, thick](0,0) -- (0,2);
		\node [draw,circle] at (-0.75,2) {$+q$};
		\node [draw,circle] at (-0.75,0) {$-q$ };
		\node [right] at (0,1) {$d$};
	\end{tikzpicture}
\end{figure}

The potential due to the dipole will be
\[
\Phi(\mathbf{x}) = \frac{q}{4 \pi \epsilon_0} \biggl( - \frac{1}{|\mathbf{x}|} + \frac{1}{|\mathbf{x} - \mathbf{d}|} \biggr)
.\]
Applying Taylor's theorem to a scalar field, we get
\[
f(\mathbf{x} + \mathbf{h}) = f(\mathbf{x}) + (\mathbf{h} \cdot \nabla) f(\mathbf{x}) + \frac{1}{2} (\mathbf{h} \cdot \nabla)^2 f(\mathbf{x}) + \mathcal{O}(|\mathbf{h}|^3)
,\]
so applying this to our potential (and letting $|\mathbf{x}| = r$,)
\begin{align*}
	\Phi(\mathbf{x}) &= \frac{q}{4 \pi \epsilon_0} \biggl( -\frac{1}{r} + \frac{1}{r} - (\mathbf{d} \cdot \nabla) \frac{1}{r} + \mathcal{O}(|\mathbf{d}|^2)\biggr) \\
			 &= \frac{q}{4 \pi \epsilon_0} \frac{\mathbf{d} \cdot \mathbf{x}}{|\mathbf{x}|^3} + \mathcal{O}(|\mathbf{d}|^2).
\end{align*}
In the limit $|\mathbf{d}| \to 0$ with $q \mathbf{d}$ finite, we obtain a \emph{point dipole}\index{point dipole} with \emph{electric dipole moment}\index{electric dipole moment}
\[
\mathbf{p} = q \mathbf{d}
,\]
with potential
\[
\Phi(\mathbf{x}) = \frac{\mathbf{x} \cdot \mathbf{x}}{4 \pi \epsilon_0 |\mathbf{x}|^3}
.\]
The electric field can be found as
\[
\mathbf{E} = - \nabla \Phi = \frac{3(\mathbf{p} \cdot \mathbf{x})\mathbf{x} - |\mathbf{x}|^3 \mathbf{p}}{4 \pi \epsilon_0 |\mathbf{x}|^{5}}
.\]
In spherical polar coordinates aligned with $\mathbf{p}= p \mathbf{e}_z$,
\begin{align*}
	\Phi &= \frac{p \cos \theta}{4 \pi \epsilon_0 r^2}, \\
	E_r &= - \frac{\partial \Phi}{\partial r} = \frac{2 p \cos \theta}{4 \pi \epsilon_0 r^3}, \\
	E_\theta &= - \frac{1}{r} \frac{\partial \Phi}{\partial \theta} = \frac{p \sin \theta}{4 \pi \epsilon_0 r^3}, \\
	E_\phi &= 0.
\end{align*}
Note that
\begin{itemize}
	\item $\Phi$ and $\mathbf{E}$ are not spherically symmetric.
	\item They decrease more rapidly with $r$ than for a point charge.
\end{itemize}

A point dipole $\mathbf{p}$ at the origin corresponds to
\begin{align*}
	\rho(\mathbf{x}) &= - \mathbf{p} \cdot \nabla \delta(\mathbf{x}), \\
	\Phi(\mathbf{x}) &= \mathbf{p} \cdot \nabla \biggl( \frac{1}{4 \pi \epsilon_0 |\mathbf{x}|} \biggr).
\end{align*}

\subsubsection{Field Lines and Equipotentials}
\label{subsubsub:field_lines_and_equipotentials}

\emph{Electric field lines}\index{electric field lines} are the integral curves of $\mathbf{E}$, being tangent to $\mathbf{E}$ everywhere.

Since $\nabla \cdot \mathbf{E} = \frac{\rho}{\epsilon_0}$, the field lines begin at positive charges and end on negative charges.

Furthermore, in electrostatics $\mathbf{E} = - \nabla \Phi$, so the field lines are perpendicular to the equipotential surface $\Phi = \rm{constant}$.


\begin{figure}[h]
	\centering
	\caption{Electric Field Lines}
	\label{fig:electric_field_lines}
	\begin{subfigure}{0.5\textwidth}
		\centering
	\begin{tikzpicture}
		\node [draw,circle] at (0,0) {$+q$};
		\draw [->](0,0.5) -- +(0,2);
		\draw [->](0.36,0.36) -- +(1.4,1.4);
		\draw [->](0.5,0) -- +(2,0);
		\draw [->](0.36,-0.36) -- +(1.4,-1.4);
		\draw [->](0,-0.5) -- +(0,-2);
		\draw [->](-0.36,-0.36) -- +(-1.4,-1.4);
		\draw [->](-0.5,0) -- +(-2,0);
		\draw [->](-0.36,0.36) -- +(-1.4,1.4);
		\draw [red] (0,0) circle (1);
		\draw [red] (0,0) circle (1.5);
		\draw [red] (0,0) circle (2);
	\end{tikzpicture}
	\caption{Point charge}
	\end{subfigure}%
	\begin{subfigure}{0.5\textwidth}
		\centering
	\begin{tikzpicture}
		\node [draw,circle] at (0,1) {$+q$};
		\node [draw,circle] at (0,-1) {$-q$};
		\draw (-3,0) -- (3,0);
		\draw [->] (0,0.5) -- +(0,-1);
		\draw [->] plot [smooth, tension = 1] coordinates {(-0.36,0.64) (-0.6, 0) (-0.36, -0.64)};
		\draw [->] plot [smooth, tension = 1] coordinates {(0.36,0.64) (0.6, 0) (0.36, -0.64)};
		\draw [->] plot [smooth, tension = 1.5] coordinates {(-0.5,1) (-1.5, 0) (-0.5, -1)};
		\draw [->] plot [smooth, tension = 1.5] coordinates {(0.5,1) (1.5, 0) (0.5, -1)};
		\draw [->] (0,1.5) -- +(0,1.1);
		\draw [<-] (0,-1.5) -- +(0,-1.1);
		\draw [red] (0,1) circle (0.7);
		\draw [red] (0,-1) circle (0.7);
		\draw [red] plot [smooth cycle, tension = 1] coordinates {(0,0.2) (1, 1) (0,2) (-1,1)};
		\draw [red] plot [smooth cycle, tension = 1] coordinates {(0,-0.2) (1, -1) (0,-2) (-1,-1)};
		\draw [red] plot [smooth cycle, tension = 1] coordinates {(0,0.1) (1.5, 1) (0,2.5) (-1.5,1)};
		\draw [red] plot [smooth cycle, tension = 1] coordinates {(0,-0.1) (1.5,-1) (0,-2.5) (-1.5,-1)};
	\end{tikzpicture}
	\caption{Dipole}
	\end{subfigure}
\end{figure}

\subsubsection{Dipole in an External Field}
\label{subsub:dipole_in_an_external_field}

Consider a dipole $\mathbf{p}$ in an external electric field $\mathbf{E} = - \nabla \Phi$ generated by distinct charges. If the dipole has charge $-q$ at $\mathbf{x}$ and $+q$ at $\mathbf{x} + \mathbf{d}$, then the potential energy of the dipole due to the external field is
\[
U = -q \Phi(\mathbf{x}) + q \Phi(\mathbf{x} + \mathbf{d}) = q (\mathbf{d} \cdot \nabla) \Phi(\mathbf{x}) + \mathcal{O}(|\mathbf{d}|^2)
.\]
In the limit of a point dipole,
\[
U = \mathbf{p} \cdot \nabla \Phi = - \mathbf{p} \cdot \mathbf{E}
.\]
This is minimized when $\mathbf{p}$ is aligned with $\mathbf{E}$.

\subsubsection{Multipole Expansion}
\label{subsub:multipole_expansion}

For a general charge distribution $\rho(\mathbf{x})$ confined to a ball $\{V \mid |\mathbf{x}| < \ell\}$, then
\[
\Phi(\mathbf{x}) = \frac{1}{4 \pi \epsilon_0} \int_{V} \frac{\rho(\mathbf{x}')}{|\mathbf{x} - \mathbf{x}'|} \Diff3 \mathbf{x}'
.\]
For an external potential with $|\mathbf{x}| > R$, we can expand
\begin{align*}
	\frac{1}{|\mathbf{x} - \mathbf{x}'|} &= \frac{1}{r} - (\mathbf{x}' \cdot \nabla) \frac{1}{r} + \frac{1}{2} (\mathbf{x}' \cdot \nabla)^2 \frac{1}{r} + \mathcal{O}(|\mathbf{x}'|^3) \\
					     &= \frac{1}{r} \biggl[ 1 + \frac{\mathbf{x}' \cdot \mathbf{x}}{r^2} + \frac{3(\mathbf{x}' \cdot \mathbf{x})^2 - |\mathbf{x}'|^2|\mathbf{x}|^2}{2r^{4}} + \mathcal{O} \biggl( \frac{R^3}{r^3} \biggr) \biggr].
\end{align*}
This leads to the \emph{multipole expansion}\index{multipole expansion} of the potential
\[
\Phi(\mathbf{x}) = \frac{1}{4 \pi \epsilon_0} \biggl( \frac{Q}{r} + \frac{\mathbf{p} \cdot \mathbf{x}}{r^3} + \frac{1}{2} \frac{Q_{ij}x_ix_j}{r^{5}} + \cdots \biggr)
.\]
The first three multipole moments are the:
\begin{itemize}
	\item total charge (or monopole moment)\index{monopole moment} - a scalar, where
		\[
			Q = \int_{V} \rho(\mathbf{x}) \Diff3 \mathbf{x}.
		\]
	\item electric dipole moment - a vector, where
		\[
			\mathbf{p} = \int_{V} \mathbf{x} \rho(\mathbf{x}) \Diff3 \mathbf{x}.
		\]
	\item electric quadrupole moment - a traceless, symmetric second order tensor
		\[
			Q_{ij} = \int_{V}(3x_ix_j - |\mathbf{x}|^2 \delta_{ij}) \rho(\mathbf{x}) \Diff3 \mathbf{x}
		\]
\end{itemize}

For $r \gg R$, $\Phi$ and $\mathbf{E}$ look increasingly like those of a point charge $Q$ unless $Q = 0$, in which case they look like those of a point dipole, unless $\mathbf{p} = 0$, etc.

\subsection{Electrostatic Energy}
\label{sub:electrostatic_energy}\index{electrostatic energy}

The work done against the electric force $\mathbf{F} = q \mathbf{E}$ in bringing a particle of charge $q$ from infinity (where we assume $\Phi = 0$) to $\mathbf{x}$ is
\[
- \int_{\infty}^{\mathbf{x}} \mathbf{F} \cdot \diff \mathbf{x} = +q \int_{\infty}^{\mathbf{x}} \nabla \Phi \cdot \diff \mathbf{x} = q \Phi (\mathbf{x})
.\]

Consider assembling a configuration of $N$ point charges one by one. Particle $i$ of charge $q_i$ is brought from $\infty$ to $\mathbf{x}_i$, while the previous particles remain fixed.
\begin{enumerate}[label = Particle \arabic*.]
	\item There is no work involved, so $W_1 = 0$.
	\item
		\[
		W_1 = q_2 \biggl( \frac{q}{4 \pi \epsilon_0 |\mathbf{x}_2 - \mathbf{x}_1|} \biggr)
		.\]
	\item
		\[
		W_3 = q_3 \biggl( \frac{q_1}{4 \pi \epsilon_0 |\mathbf{x}_3 - \mathbf{x}_1|} + \frac{q_2}{4 \pi \epsilon_0|\mathbf{x}_3 - \mathbf{x}_2|} \biggr)
		,\]
\end{enumerate}
and so on. The total work done is
\[
U = \sum_{i = 1}^{N} W_i = \sum_{i = 2}^{N} \sum_{j = 1}^{i-1} \frac{q_i q_j}{4 \pi \epsilon_0 |\mathbf{x}_i - \mathbf{x}_j|}
.\] 
This can be rewritten as
\[
	U = \frac{1}{2} \sum_{i = 1}^{N}\sum_{\substack{j=1\\j\neq i}}^{N} \frac{q_i q_j}{4 \pi \epsilon_0|\mathbf{x}_i - \mathbf{x}_j|}
,\]
or
\[
U = \frac{1}{2} \sum_{i = 1}^{N} q_i \Phi(\mathbf{x}_i)
.\]

Generalizing to a continuous charge distribution $\rho(\mathbf{x})$, occupying a finite volume $V$,
\[
U = \frac{1}{2} \int_{V} \rho(\mathbf{x}) \Phi(\mathbf{x}) \Diff3 \mathbf{x} = \frac{1}{2} \int_{V} \rho \Phi \diff V
.\]
Using (\ref{eq:m1}) we have
\begin{align*}
	U &= \frac{1}{2} \int_{V} (\epsilon_0 \nabla \cdot \mathbf{E}) \Phi \diff V = \frac{\epsilon_0}{2} \int_{V} (\nabla \cdot (\Phi \mathbf{E}) - \mathbf{E} \cdot \nabla \Phi) \diff V \\
	  &= \frac{\epsilon_0}{2} \int_{S} \phi \mathbf{E} \cdot \diff \mathbf{S} + \int_{V} \frac{\epsilon_0|\mathbf{E}|^2}{2} \diff V
.\end{align*}

Let $S = \partial V$ be a sphere of radius $R \to \infty$. Then $\Phi = \mathcal{O}(R^{-1})$, and $\mathbf{E} = \mathcal{O}(R^{-2})$ on $S$, while the area of $S$ is $\mathcal{O}(R^2)$, so the area integral is $\mathcal{O}(R^{-1})$ and goes to zero as $R \to \infty$. Thus,
\[
U = \int \frac{\epsilon_0 |\mathbf{E}|^2}{2} \diff V
,\]
integrated over all space.

This implies that energy is stored in the electric field, even in a vacuum.

Any of the expression for $U$ suggest that the self-energy of a point charge is infinite. We can discard this as it is unchanging and causes no force.

\subsection{Conductors}
\label{sub:conductors}

In an \emph{conductor}\index{conductor} such as a metal, some charges (usually electrons) can move freely. In electrostatics we require
\[
	\mathbf{E} = \mathbf{0}, \qquad \Phi = \rm{constant}
\]
inside a conductor, hence $\rho = 0$. Otherwise free charges would move in response to the electric force and a current would flow.

A surface charge density $\rho$ can exist on the surface of a conductor, which is an equipotential.

Taking a normal $\mathbf{n}$ to the point of the conductor, the condition
\[
	[\mathbf{n} \cdot \mathbf{E}] = \frac{\sigma}{\epsilon_0} \implies \mathbf{n} \cdot \mathbf{E} = \frac{\sigma}{\epsilon_0}
\]
immediately outside the conductor.

The constant potential of a conductor can be set by connecting it to a battery or another conductor. An \emph{earthed} (or \emph{grounded}) conductor is connected to the ground, usually taken as $\Phi = 0$.

To find $\Phi(\mathbf{x})$ and $\mathbf{E}(\mathbf{x})$ due to a charge distribution $\rho(\mathbf{x})$ in the presence of conductors with surfaces $S_i$ and potentials $\Phi_i$, we solve Poisson's equation
\[
- \nabla^2 \Phi = \frac{\rho}{\epsilon_0}
,\]
with Dirichlet boundary conditions $\Phi = \Phi_i$ on $S_i$. The solution depends linearly on $\rho$ and $\{\Phi_i\}$.

\begin{exbox}
	Consider a point charge $q$ at position $(0,0,h)$ in a half-space $z > 0$, bounded by an earthed conducting wall ($\Phi = 0$ on $z = 0$).

	By the method of images, the solution in $z > 0$, is identical to that of a dipole, with image charge $-q$ at $(0, 0, -h)$.

	This is as the wall coincides with an equipotential of the dipole. The induced surface charge density on the wall can be worked out from
	\[
	\frac{\sigma}{\epsilon_0} = \mathbf{n} \cdot \mathbf{E} = E_z = - \frac{qh}{4 \pi \epsilon_0 (r^2 + h^2)^{3/2}}
	,\]
	where $r = \sqrt{x^2 + y^2}$. The total induced surface charge is
	\[
	\int_{0}^{\infty} \sigma 2 \pi r \diff r = - q h \int_{0}^{\infty} \frac{r \diff r}{(r^2 + h^2)^{3/2}} = -q
	.\]
\end{exbox}

\begin{figure}[h]
	\centering
	\caption{Point Charge and Wall}
	\label{fig:point_charge_and_wall}
	\begin{tikzpicture}
		\draw[fill] (0,2) circle (0.05);
		\draw[fill] (0,-2) circle (0.05);
		\draw (-5,0) -- (5,0);
		\draw[<->] (0,0.1) -- (0,1.85);
		\draw[<->] (0,-0.1) -- (0,-1.85);
		\node[right] at (0,1) {$h$};
		\node[right] at (0,-1) {$h$};
		\node[right] at (0,2) {$+q$};
		\node[right] at (0,-2) {$-q$};
		\draw (0.15, 1.85) -- (1.85,0.1);
		\node[below] at (0.925,0) {$r$};
		\draw[->] (2.5,0.1) -- (2.5,0.8);
		\node[above] at (2.5,0.8) {\small$z>0$};
		\node[right] at (5,0) {\small$z=0$};
	\end{tikzpicture}
\end{figure}

A simple \emph{capacitor}\index{capacitor} consists of two separated conductors carrying charges $\pm Q$.

If the potential difference (voltage) between them is $V$, then the capacitance\index{capacitance} is defined by
\[
C = \frac{Q}{V}
,\]
and depends only on the geometry, because $\Phi$ depends linearly on $Q$.

\begin{exbox}
	Consider two infinite parallel plates separated by $d$. Let the plate surfaces be at $z = 0$, $z = d$, and have surface charge densities $\pm \sigma$. Then, $\mathbf{E} = E \mathbf{e}_z$ with $E = \sigma/\epsilon_0$ constant for $0 < z < d$.

	Then $\Phi = - Ez + \rm{constant}$ and $V = Ed$.

	The same solution holds approximately for parallel plates of area $A \gg d^2$ if end-effects are neglected. So,
	\[
	C = \frac{Q}{V} \approx \frac{\sigma A}{E d} \approx \frac{\epsilon_0 A}{d}
	.\]
	The electrostatic energy stored in the capacitor is
	\[
	U = \int \frac{\epsilon_0 |\mathbf{E}|^2}{2} \diff V \approx \frac{\epsilon_0 E^2}{2} A d \approx \frac{1}{2} CV^2
	.\]
	In general,
	\[
	U = \frac{1}{2} CV^2 = \frac{Q^2}{2C}
	.\]
	The work done in moving an element of charge $\delta Q$ from one plate to another is $\delta W = V \delta Q$. So the total work done is
	\[
	\int_{0}^{Q} \frac{Q'}{C} \diff Q' = \frac{Q^2}{2C}
	.\]
	Or we can use
	\[
	U = \frac{1}{2} \int \rho \Phi \diff V = \frac{1}{2} Q \Phi_{+} - \frac{1}{2} Q \Phi_{-} = \frac{1}{2} Q V
	.\]
\end{exbox}

\begin{figure}[h]
	\centering
	\caption{Capacitors}
	\label{fig:capacitors}
	\begin{tikzpicture}
		\draw (-5, -2) -- (5, -2);
		\draw (-5, 2) -- (5, 2);
		\draw[->] (-3, -1) -- (-3, 1);
		\draw[->] (-1, -1) -- (-1, 1);
		\draw[->] (1, -1) -- (1, 1);
		\draw[->] (3, -1) -- (3, 1);
		\node[left] at (-5,-2) {$+\sigma$};
		\node[left] at (-5,2) {$-\sigma$};
		\node[right] at (5,-2) {$z=0$};
		\node[right] at (5,2) {$z=d$};
		\fill[pattern = north west lines] (-5,2) rectangle +(10,0.5);
		\fill[pattern = north east lines] (-5,-2) rectangle +(10,-0.5);
		\node at (0,0) {$\mathbf{E}$};
	\end{tikzpicture}
\end{figure}

\newpage

\section{Magnetostatics}
\label{sec:magnetostatics}

\emph{Magnetostatics}\index{magnetostatics} is the study of the magnetic field generated by a stationary current distribution:
\begin{align*}
	\nabla \times \mathbf{B} &= \mu_0 \mathbf{J} \tag{M4'}\label{eq:m4'}\\
	\nabla \cdot \mathbf{B} &= 0 \tag{M2}\label{eq:m2}
\end{align*}

From (\ref{eq:m4'}), we get $\nabla \cdot \mathbf{J} = 0$, the time-independent equation of charge conservation.

\subsection{Amp\`{e}re's Law}
\label{sub:amperes_law}

Consider a closed curve $C$ that is the boundary of an open surface $S$. Integrate (\ref{eq:m4'}) over $S$ and applying Stokes' theorem, we obtain \emph{Amp\`{e}re's law}\index{Amp\`{e}re's law}
\[
\int_{C} \mathbf{B} \cdot \diff \mathbf{x} = \mu_0 I
,\]
where
\[
I = \int_{S} \mathbf{J} \cdot \diff \mathbf{S}
\]
is the total current through $S$.

Since $\nabla \cdot \mathbf{J} = 0$, the same current $I$ flows through any open surface $S$ such that $\partial S = C$.

Amp\`{e}re's law is the integral version of (\ref{eq:m4'}) and is valid provided $\mathbf{E}$ is constant through time. In words, it says:
\begin{center}
	The circulation of magnetic field around a loop is proportional to the total current through the loop.
\end{center}

In special situations, we can use Amp\`{e}re's law, together with symmetry to deduce $\mathbf{B}$ from $\mathbf{J}$.

A cylindrically symmetric situation could involve:
\begin{itemize}
	\item An axial current distribution $J_z(r)\mathbf{e}_z$,
	\item An azimuthal current distribution $J_{\phi}(r)\mathbf{e}_{\phi}$,
\end{itemize}
or a combination. Since $\nabla \cdot \mathbf{J} = 0$, we have no radial component.

The same applies to $\mathbf{B}$. Hence the curl in Maxwell's equations implies $B_{\phi}$ is linearly related to $J_z$, and $B_z$ is linearly related to $J_{\phi}$.

\subsubsection{Long Straight Wire}
\label{subsub:long_straight_wire}

A cylindrical wire of radius $R$ carries a total current $I$ parallel to its axis.

To find $B_{\phi}(r)$ generated by $J_z(r)$, we apply Amp\`{e}re's law to a circle $C$ of radius $r$. Here $S$ is a disc.

\begin{figure}[h]
	\centering
	\caption{Long Straight Wire}
	\label{fig:long_straight_wire}
	\begin{tikzpicture}
		\draw (-1, -3) -- (-1, 3);
		\draw (1, -3) -- (1, 3);
		\draw (0,3) circle (1 and 0.3);
		\draw (-1,-3) arc (180:360:1 and 0.3);
		\draw[red] (-1,1) arc (113.5:426.5:2.5 and 0.75);
		\draw (0,3) -- (0.5,3.253);
		\node[below] at (0.5, 3.253) {\small $R$};
		\draw[->] (0,1.7) -- (0, 2.1);
		\node at (0, 2.3) {\small $\mathbf{I}$};
		\draw[red] (0,0.35) -- (1.5,0.92);
		\node[red] at (0.5, 0.3) {\small $r$};
		\node[red] at (-0.5, 0.35) {\small $S$};
	\end{tikzpicture}
\end{figure}

If $r > R$, then
\begin{align*}
	\int_{C} \mathbf{B} \cdot \diff \mathbf{x} &+ B_{\phi}(r) \int_{C} \mathbf{e}_{\phi} \cdot \diff \mathbf{x} = B_{\phi}(r) \int_{C} \diff \ell \\
						   &= B_{\phi}(r) 2 \pi r = \mu_0 I.
\end{align*}

Therefore, outside the wire,
\[
\mathbf{B} = \frac{\mu_0 I}{2 \pi r} \mathbf{e}_{\phi}
.\]

\subsubsection{Solenoid}
\label{subsub:solenoid}

A thin wire is coiled arund a cylindrical tube of radius $R$. An \emph{ideal solenoid}\index{ideal solenoid} is infinitely long and tightly wound, having cylindrical geometry and purely azimuthal current.

The wire carries current $I$ and has $N$ turns per unit length of the tube.

\begin{figure}[h]
	\centering
	\caption{Solenoid}
	\label{fig:solenoid}
	\begin{tikzpicture}
		\draw (-1, -3) -- (-1, 3);
		\draw (1, -3) -- (1, 3);
		\draw (0,3) circle (1 and 0.3);
		\draw (-1,-3) arc (180:360:1 and 0.3);
		\draw (0,3) -- (0.5,3.253);
		\node[below] at (0.5, 3.253) {\small $R$};
		\draw[blue,middlearrow={>}] (-1,-2.2) to[out=170,in=-10] (1,-1.9);
		\draw[blue,middlearrow={>}] (-1,-2) to[out=170,in=-10] (1,-1.7);
		\draw[blue,middlearrow={>}] (-1,-1.8) to[out=170,in=-10] (1,-1.5);
		\draw[blue,middlearrow={>}] (-1,-1.6) to[out=170,in=-10] (1,-1.3);
		\draw[blue,middlearrow={>}] (-1,-1.4) to[out=170,in=-10] (1,-1.1);
		\draw[blue,middlearrow={>}] (-1,-1.2) to[out=170,in=-10] (1,-0.9);
		\draw[blue,middlearrow={>}] (-1,-1) to[out=170,in=-10] (1,-0.7);
		\draw[red,middlearrow={>}] (0.7, -1.9) -- (0.7,-0.9);
		\draw[red,middlearrow={>}] (0.7, -0.9) -- (1.7,-0.9);
		\draw[red,middlearrow={>}] (1.7,-0.9) -- (1.7,-1.9);
		\draw[red,middlearrow={>}] (1.7,-1.9) -- (0.7,-1.9);
		\node[red] at (1.4, -1.4) {\small $S$};
		\node[red] at (1.4,-0.7) {\small $C$};
		\draw[red,|-|] (0,-0.6) -- (0.7,-0.6);
		\node[red,above] at (0.35,-0.6) {$a$};
		\draw[red,|-|] (0, -0.1) -- (1.7, -0.1);
		\node[red,above] at (0.85,-0.1) {$b$};
		\draw[red,|-|] (2,-1.9) -- (2,-0.9);
		\node[red,right] at (2,-1.45) {$L$};
	\end{tikzpicture}
\end{figure}

To find $B_z(r)$ generated by $J_{\phi}(r)$, we apply Amp\`{e}re's law to a rectangular loop $C$. Taking $a < b < R$ or $R < a < b$ gives
\[
L(B_z(a) - B_z(b)) = 0
.\]
Taking $a < R < b$ gives
\[
L(B_z(a) - B_z(b)) = \mu_0 N L I
.\]
Assuming that $B_z(r) \to 0$ as $r \to \infty$, we deduce that
\[
B_z(r) =
\begin{cases}
	\mu_0 N I & r < R,\\
	0 & r > R.
\end{cases}
\]

The ideal solenoid is an example of a \emph{surface current}\index{surface current}. Here it is of the form
\[
J_{\phi}(r) = K_{\phi}\delta(r - R),
.\]
where $K_{\phi} = NI$. Generally, a \emph{surface current density}\index{surface current density} $\mathbf{K}$ produces a discontinuity in the tangential magnetic field:
\[
	[\mathbf{n} \times \mathbf{B}] = \mu_0 \mathbf{K}
.\]
This follows from Ampere's law applied to a loop through a surface, where we take $L, h \to 0$.

\begin{figure}[h]
	\centering
	\caption{Surface Current}
	\label{fig:surface_current}
	\begin{tikzpicture}
		\draw (-7,0) -- (1,-1) -- (7,1) -- (-1,2) -- cycle;
		\draw[red] (0,0) -- (0,0.2) -- (1,0.5333) -- (1,0.3333);
		\draw[red,dashed] (1,0.3333) -- (1,0) -- (0,-0.3333) -- (0,0);
		\draw[->] (4, 0.5) -- (4,0.9);
		\node[above] at (4,0.9) {\small $\mathbf{n}$};
		\draw[red,|-|] (0,0.5) -- (1,0.8333);
		\node[red,above] at (0.5,0.7) {\small $L$};
		\draw[red,|-|] (1.2,0.6333) -- (1.2,0.3);
		\node[red,right] at (1.2,0.5) {\small $h$};
	\end{tikzpicture}
\end{figure}

Applying the same reasoning with (\ref{eq:m2}), we get
\[
	[\mathbf{n} \cdot \mathbf{B}] = 0
,\]
so the normal component is continuous.

\subsection{Magnetic Vector Potential}
\label{sub:magnetic_vector_potential}

(\ref{eq:m2}) implies that $\mathbf{B}$ can be written in terms of a \emph{magnetic vector potential}\index{magnetic vector potential} $\mathbf{A}(\mathbf{x})$:
\[
\mathbf{B} = \nabla \times \mathbf{A}
.\]
$\mathbf{A}$ is not unique. If we make a \emph{gauge transformation}\index{gauge transformation}, replacing $\mathbf{A}$ with
\[
\mathbf{A}' = \mathbf{A} + \nabla \chi
,\]
where $\chi(\mathbf{x})$ is an arbitrary scalar field, then $\mathbf{B}$ is unchanged, as
\[
\mathbf{B} = \nabla \times \mathbf{A} = \nabla \times \mathbf{A}'
.\]
A convenient gauge for many calculation is the \emph{Coulomb gauge}\index{Coulomb gauge} in which $\nabla \cdot \mathbf{A} = 0$.

We can assume this condition without loss of generality. If $\nabla \cdot \mathbf{A} \neq 0$, then we can make a gauge transformation $\nabla \cdot \mathbf{A}' = 0$ by choosing $\chi$ to be the solution of Poisson's equation
\[
- \nabla^2 \chi = \nabla \cdot \mathbf{A}
.\]
In terms of $\mathbf{A}$, (\ref{eq:m4'}) becomes
\[
\nabla \times (\nabla \times \mathbf{A}) = \mu_0 \mathbf{J}
.\]
Using the identity
\[
	\nabla \times (\nabla \times \mathbf{A}) = \nabla (\nabla \cdot \mathbf{A}) - \nabla^2 \mathbf{A}
,\]
and assuming a Coulomb gauge, we obtain Poisson's equation in vector form:
\[
- \nabla^2 \mathbf{A} = \mu_0 \mathbf{J}
.\]

\subsection{The Biot-Savart Law}
\label{sub:the_biot_savart_law}

The solution of Poisson's equation is
\[
\mathbf{A}(\mathbf{x}) = \frac{\mu_0}{4 \pi} \int \frac{\mathbf{J}(\mathbf{x}')}{|\mathbf{x} - \mathbf{x}'|} \Diff3 \mathbf{x}'
.\]
We should check that the solution satisfies the assumed Coulomb gauge condition:
\begin{align*}
	\nabla \cdot \mathbf{A}(\mathbf{x}) &= \frac{\mu_0}{4 \pi} \int_{V} \nabla \cdot \biggl( \frac{\mathbf{J}(\mathbf{x}')}{|\mathbf{x} - \mathbf{x}'|} \biggr) \Diff3 \mathbf{x}' \\
					    &= \frac{\mu_0}{4 \pi} \int_{V} \mathbf{J}(\mathbf{x}') \cdot \nabla \biggl( \frac{1}{|\mathbf{x} - \mathbf{x}'|} \biggr) \Diff3 \mathbf{x}' \\
					    &= - \frac{\mu_0}{4 \pi} \int_{V} \mathbf{J}(\mathbf{x}') \cdot \nabla' \biggl( \frac{1}{|\mathbf{x} - \mathbf{x}'|} \biggr) \Diff3 \mathbf{x}' \\
					    &= - \frac{\mu_0}{4 \pi} \int_{V} \nabla' \cdot \biggl( \frac{\mathbf{J}(\mathbf{x}')}{|\mathbf{x} - \mathbf{x}'|} \biggr) \Diff3 \mathbf{x}' \\
					    &= - \frac{\mu_0}{4 \pi} \int_{\partial V} \frac{\mathbf{J}(\mathbf{x}') \cdot \diff \mathbf{S}'}{|\mathbf{x} - \mathbf{x}'|}.
\end{align*}
This is $0$, as assumed, if the current is contained in some finite volume and we take $V$ to be at least as large, or if $\mathbf{J}$ decays sufficiently as $|\mathbf{x}| \to \infty$.

To find the magnetic field, derive $\mathbf{B} = \nabla \times \mathbf{A}$ to get
\[
\mathbf{B}(\mathbf{x}) = \frac{\mu_0}{4 \pi} \int \frac{\mathbf{J}(\mathbf{x}') \times (\mathbf{x} - \mathbf{x}')}{|\mathbf{x} - \mathbf{x}'|^3} \Diff3 \mathbf{x}'
.\]
This is the \emph{Biot-Savart law}\index{Biot-Savart law}, giving the magnetic field generated by a stationary current distribution.

A special case is when the current is restricted to a thin wire in the form of a curve $C$. Then the courrent element $\mathbf{J} \Diff3 \mathbf{x}$ can be replaced by $I \diff \mathbf{x}$. Charge conservation means that $I$ is constant along the wire, so
\[
\mathbf{B}(\mathbf{x}) = \frac{\mu_0 I}{4 \pi} \int_{C} \frac{\diff \mathbf{x}' \times (\mathbf{x} - \mathbf{x}')}{|\mathbf{x} - \mathbf{x}'|^3}
.\]
Another way to derive this is using delta functions. The thin wire current density can be represented as
\[
\mathbf{J}(\mathbf{x}) = I \int_{C} \delta(\mathbf{x} - \mathbf{x}') \diff \mathbf{x}'
.\]
Substituting this into the Biot-Savart law, gives the same result. Note that charge conservation takes the form
\begin{align*}
	\nabla \cdot \mathbf{J}(\mathbf{x}) &= I \int_{C} \nabla \delta(\mathbf{x} - \mathbf{x}') \cdot \diff \mathbf{x}' \\
					    &= -I \int_{C} \nabla' \delta(\mathbf{x} - \mathbf{x}') \cdot \diff \mathbf{x}' \\
					    &= -I [\delta(\mathbf{x} - \mathbf{x}_2) - \delta(\mathbf{x} - \mathbf{x}_1)],
\end{align*}
where $C$ runs from $\mathbf{x}_1$ to $\mathbf{x}_2$. If $C$ is closed then $\mathbf{x}_2 = \mathbf{x}_1$, and $\nabla \cdot \mathbf{J} = 0$ as expected. If $C$ is infinite, then $\nabla \cdot \mathbf{J} = 0$ for any finite $\mathbf{x}$.

We can check that the Biot-Savart law gives the same result as Amp\`{e}re's law for a long straight thin wire:

\begin{figure}[h]
	\centering
	\caption{Thin Wire Magnetic Field}
	\label{fig:twmf}
	\begin{tikzpicture}
		\draw[->] (0,-3) -- (0,3);
		\draw (0,3) -- (0,4);
		\draw[fill] (3,-1) circle (0.05);
		\draw (3,-1) -- (0,2);
		\draw (3,-1) -- (0,-1);
		\node[below] at (1.5,-1) {\small $r$};
		\node[right] at (3,-1) {\small $P$};
		\node[left] at (0,-1) {\small $0$};
		\node[left] at (0,3) {\small $I$};
		\node[left] at (0, 0.5) {\small $z'$};
	\end{tikzpicture}
\end{figure}

We have $\mathbf{x} = r \mathbf{e}_r$, taking $z = 0$ by translation symmetry, and $\mathbf{x}' = z' \mathbf{e}_z$. Hence $\mathbf{x} - \mathbf{x}' = r \mathbf{e}_r - z' \mathbf{e}_z$, and $\diff \mathbf{x}' = \diff z' \mathbf{e}_z$, giving
\begin{align*}
	\mathbf{B}(\mathbf{x}) &= \frac{\mu_0 I}{4 \pi} \mathbf{e}_{\phi} \int_{-\infty}^{\infty} \frac{r \diff z'}{(r^2 + z'^2)^{3/2}} \\
			       &= \frac{\mu_0 I}{4 \pi} \mathbf{e}_{\phi} \biggl[ \frac{z'}{r(r^2 + z'^2)^{1/2}} \biggr]_{-\infty}^{\infty} \\
			       &= \frac{\mu_0 I}{2 \pi r} \mathbf{e}_{\phi}.
\end{align*} 

\subsection{Magnetic Dipole}
\label{sub:magnetic_dipole}

For a general current distribution $\mathbf{J}(\mathbf{x})$ confined to a ball $\{V \mid |\mathbf{x}| < R\}$,
\[
\mathbf{A}(\mathbf{x}) = \frac{\mu_0}{4 \pi} \int_{V} \frac{\mathbf{J}(\mathbf{x'})}{|\mathbf{x} - \mathbf{x}'|} \Diff3 \mathbf{x}'
.\]
The external field for $|\mathbf{x}| = r > R$ can be evaluated by expanding
\[
\frac{1}{|\mathbf{x} - \mathbf{x}'|} = \frac{1}{r} \biggl( 1 + \frac{\mathbf{x}' \cdot \mathbf{x}}{r^2} + \mathcal{O}\biggl(\frac{R^2}{r^2} \biggr) \biggr)
,\]
leading to a multipole expansion, as for the electric field. To do this, we need to calculate the moments of the current distribution.

Since $\mathbf{J} = \mathbf{0}$ on $\partial V$ and $\nabla \cdot \mathbf{J} = 0$, the divergence theorem implies
\begin{align*}
	0 &= \int_{\partial V} x_i J_j \diff S_j = \int_{V} \partial_j (x_i J_j) \diff3 \mathbf{x} \\
	  &= \int_{V} (\delta_{ij} J_j + x_j \partial_j J_j) \Diff3 \mathbf{x} \\
	  &= \int_{V} J_i \Diff3 \mathbf{x}
.\end{align*}
So the zeroth moment vanishes. Similarly,
\begin{align*}
	0 &= \int_{\partial V} x_i x_j J_k \diff S_k = \int_{V} \partial_k (x_i x_j J_k) \Diff3 \mathbf{x} \\
	  &= \int_{V}(\delta_{ik} x_j J_k + x_j \delta_{jk} J_k + x_i x_j \partial_k J_k) \Diff3 \mathbf{x} \\
	  &= \int_{V} x_j J_i \Diff3 \mathbf{x} + \int_{V} x_i J_j \Diff3 \mathbf{x}.
\end{align*}
The first moment is an antisymmetric matrix. The \emph{magnetic dipole moment}\index{magnetic dipole moment} is
\[
	\mathbf{m} = \frac{1}{2} \int_{V} \mathbf{x} \times \mathbf{J} \Diff3 \mathbf{x}
,\]
so
\[
m_i = \frac{1}{2} \epsilon_{ijk} \int_{V} x_j J_k \Diff3 \mathbf{x}
.\]
This is a vector related to the antisymmetric matrix by
\[
\int_{V} x_i J_j \Diff3 \mathbf{x} = \epsilon_{ijk} m_k
.\]
Returning to the multipole expansion for $\mathbf{A}$, we have
\begin{align*}
	A_i(\mathbf{x}) &= \frac{\mu_0}{4 \pi |\mathbf{x}|} \Biggl( \int_{V} J_i(\mathbf{x}') \Diff3 \mathbf{x}'+ \frac{x_j}{|\mathbf{x}|^3} \int_{V} x_j' J_i(\mathbf{x}') \Diff3 \mathbf{x}' + \cdots \Biggr) \\
	&= \frac{\mu_0}{4 \pi|\mathbf{x}|} \biggl( 0 + \frac{x_j \epsilon_{jik} m_k}{|\mathbf{x}|^3} + \cdot \biggr).
\end{align*}

The leading approximation is therefore
\[
	\mathbf{A}(\mathbf{x}) = \mathbf{A}_{\mathrm{dipole}}(\mathbf{x}) = \frac{\mu_0}{4 \pi} \frac{\mathbf{m} \times \mathbf{x}}{|\mathbf{x}|^3}
.\]
which is the vector potential due to a point dipole $\mathbf{m}$ at the origin. The corresponding magnetic field is
\[
	\mathbf{B}_{\mathrm{dipole}} = \nabla \times \mathbf{A}_{\mathrm{dipole}} = \frac{\mu_0}{4 \pi} \biggl( \frac{3(\mathbf{x} \cdot \mathbf{x})\mathbf{x} - |\mathbf{x}|^3 \mathbf{m}}{|\mathbf{x}|^{5}} \biggr)
.\]
A point dipole $\mathbf{m}$ at the origin corresponds to the current density and vector potential
\[
\mathbf{J} = \nabla \times (\mathbf{m} \delta(\mathbf{x})), \qquad \mathbf{A} = \nabla \times \biggl( \frac{\mu_0 \mathbf{m}}{4 \pi |\mathbf{x}|} \biggr)
.\]
The magnetic dipole moment of a thin wire carrying current $I$ around a closed curve $C$ is
\[
\mathbf{m} = \frac{I}{2} \int_{C} \mathbf{x} \times \diff \mathbf{x}
.\]
To evaluate this, let $\mathbf{a}$ be any constant vector. Then by Stokes' theorem,

\begin{align*}
	\mathbf{a} \cdot \mathbf{m} &= \frac{I}{2} \int_{C} \mathbf{a} \cdot (\mathbf{x} \cdot \diff \mathbf{x}) = \frac{I}{2} \int_{C} (\mathbf{a} \times \mathbf{x}) \cdot \diff \mathbf{x} \\
				    &= \frac{I}{2} \int_{S}(\nabla \times(\mathbf{a} \times \mathbf{x})) \cdot \diff \mathbf{S} = I \int_{S} \mathbf{a} \cdot \diff \mathbf{S},
\end{align*}
where $S$ is an open surface with boundary $C$, and we use
\begin{align*}
	\nabla \times (\mathbf{a} \times \mathbf{x}) &= \mathbf{x} \cdot \nabla \mathbf{a} - \mathbf{a} \cdot \nabla \mathbf{x} + (\nabla \times \mathbf{x})\mathbf{a} - (\nabla \times \mathbf{a})\mathbf{x} \\
						     &= \mathbf{0} - \mathbf{a} + 3 \mathbf{a} - \mathbf{0} = 2 \mathbf{a}.
\end{align*}
Since $\mathbf{a}$ is arbitrary, we obtain
\[
\mathbf{m} = I \mathbf{S}
,\]
where
 \[
\mathbf{S} = \int_{S} \diff \mathbf{S}
\]
is the vector area of $S$, which depends only on $C$, not on the choice of $S$.

\begin{exbox}
	Consider a circular loop with $x^2 + y^2 = a^2$, $\tau = 0$, for which $\mathbf{m} = I \pi a^2 \mathbf{e}_z$.

	On the $z$-axis, the dipole approximation gives
	\[
	B_z = \frac{\mu_0}{4 \pi} \biggl( \frac{3m_z z^2 - z^2 m_z}{|z|^{5}} \biggr) = \frac{\mu_0 I a^2}{2|z|^3}
	.\]
	The exact solution is
	\[
		B_z = \frac{\mu I a^2}{2(z^2 + a^2)^{3/2}}
	.\]
\end{exbox}

%See EM 9/2/23 for magnetic field lines for a current around a wire.

\subsection{Permanent Magnets}
\label{sub:permanent_magnets}

A bar magnet has north and south poles and a dipole moment. This comes from the superposition of aligned dipoles on the atomic scale. Atoms contain electrons, which are spinning charged particles, with magnetic dipole moment.

A classical model of a particle is a spinning charged sphere, which is a current loop with a magnetic dipole moment proportional to its charge and spin.

As far as we know, there are no magnetic charges (monopoles).

The Earth may also be viewed as a magnet. The liquid iron outer core of the Earth is a conducting fluid in convective motion and supports electric currents that generate a magnetic field. At the Earth's surface, this resembles a dipole field.

\subsection{Magnetic Forces}
\label{sub:magnetic_forces}

The \emph{Lorentz force}\index{Lorentz force} on a particle of charge $q$ at position $\mathbf{x}_i(t)$ is
\[
q(\mathbf{E} + \mathbf{\dot x}_i \times \mathbf{B})
.\]
In the limit of continuous charge and current densities, the Lorentz force per unit volume is then
\[
\rho \mathbf{E} + \mathbf{J} \times \mathbf{B}
.\]
We can recover the discrete version by substituting
\begin{align*}
	\rho &= \sum_{i} q_i \delta(\mathbf{x} - \mathbf{x}_i(t)), \\
	\mathbf{J} &= \sum_{i} q-i \mathbf{\dot x}_i (t) \delta(\mathbf{x} - \mathbf{x}_i(t)).
\end{align*}

Consider two or more think wires with currents $I_i$ along curves $C_i$. The total magnetic field $\mathbf{B}= \sum_i \mathbf{B}_i$, where
\[
\mathbf{B}_i(\mathbf{x}) = \frac{\mu_0 I_i}{4 \pi} \int_{C_i} \frac{\diff \mathbf{x}_i \times (\mathbf{x} - \mathbf{x}_i)}{|\mathbf{x} - \mathbf{x}_i|^3}
\]
is the magnetic field due to wire $i$. The current density is $\mathbf{J} = \sum_i \mathbf{J}_i$, where
\[
\mathbf{J}_i(\mathbf{x}) = I_i \int_{C_i}\delta(\mathbf{x} - \mathbf{x}_i) \diff \mathbf{x}_i
.\]
The total magnetic force acting on a volume $V$ is
\[
\mathbf{F} = \int_{V} \mathbf{J} \times \mathbf{B} \diff V
.\]
The force acting on wire $i$ is
\[
\mathbf{F}-i = \int \mathbf{J}_i(\mathbf{x}) \times \mathbf{B}(\mathbf{x}) \Diff3 \mathbf{x} = I_i \int_{C_i} \diff \mathbf{x}_i \times \mathbf{B}(\mathbf{x}_i)
.\]
Since $\mathbf{B} = \sum_i \mathbf{B}_i$, we have
\[
\mathbf{F}_i = \sum_{j} \mathbf{F}_{ij}
,\]
where
\[
\mathbf{F}_{ij} = I_i \int_{C_i} \diff \mathbf{x}_i \times \mathbf{B}_j(\mathbf{x}_i)
\]
is the force on wire $i$ due to wire $j$. Using the Biot-Savart law,
\[
\mathbf{F}_{ij} = \frac{\mu_0 I_i I_j}{4 \pi} \int_{C_i} \int_{C_j} \diff \mathbf{x}_i \times \biggl( \frac{\diff \mathbf{x}_j \times (\mathbf{x}_i - \mathbf{x}_j)}{|\mathbf{x}_i - \mathbf{x}_j|^3} \biggr)
.\]
This can be rewritten in a manifestly antisymmetric way that shows that
\[
\mathbf{F}_{ji} = - \mathbf{F}_{ij}
,\]
as expected from Newton's third law. The self force $\mathbf{F}_{ii}$ vanishes, although the thin-wire integral is singular, and it is better to treat the case of thick wires.

Consider two infinitely long, parallel, thin wires separated by a distance $r$. Use cylindrical polars centred on wire two, we have
\[
\mathbf{B}_2 = \frac{\mu_0 I_2}{2 \pi r} \mathbf{e}_{\phi}, \qquad \mathbf{F}_{12} = I_u \int_{-\infty}^{\infty} \diff z \mathbf{e}_2 \times \mathbf{B}_2
.\]
The total force is infinite. The force per unit length is
\[
I \mathbf{e}_z \times \mathbf{B}_z = - \frac{\mu_0 I_1 I_2}{2 \pi r} \mathbf{e}_r
.\]
This is directed towards wire two if $I_1I_2 > 0$. So the force is attractive if the currents are aligned, and repulsive otherwise.

\subsection{Force and Torque on a Magnetic Dipole}

Consider a localized current distribution confined to a ball $\{V \mid |\mathbf{x}| < R\}$. Place this in an external magnetic field $\mathbf{B}(\mathbf{x})$ that varies slowly over the length scale $R$.

The magnetic torque (about the origin) on the current loop is
\begin{align*}
	\bm{\tau} &= \int_{V} \mathbf{x} \times (\mathbf{J} (\mathbf{x}) \times \mathbf{B}(\mathbf{x})) \Diff3 \mathbf{x} \\
		  &= \int_{V}((\mathbf{x} \cdot \mathbf{B}(\mathbf{x})) \mathbf{J}(\mathbf{x}) - (\mathbf{x} \cdot \mathbf{J}(\mathbf{x}))\mathbf{B}(\mathbf{x})) \Diff3 \mathbf{x}.
\end{align*}
Within $V$, $\mathbf{B}(\mathbf{x})$ can be expressed as a Taylor series
\[
B_i(\mathbf{x})= B_i(\mathbf{0}) + x_j \partial_j B_i(\mathbf{0}) + \cdots
\]
Retaining only the zeroth-order term, we have
\[
\tau_i \approx B_j(\mathbf{0}) \int_{V} x_j J_i \Diff3 \mathbf{x} - B_i(\mathbf{0}) \int_{V} x_j J_j \Diff3 \mathbf{x}
.\]
Recall the first moments of the current distribution
\[
\int_{V} x_i J_j \Diff3 \mathbf{x} = \epsilon_{ijk} m_k
.\]
Thus $\tau_i \approx B_j(\mathbf{0}) \epsilon_{jik}m_k$. In general,
\[
\bm{\tau} \approx \mathbf{m} \times \mathbf{B}
.\]
For the force, we need to go to the first order of the Taylor expansion of $\mathbf{B}$:
\begin{align*}
	\mathbf{F} &= \int_{V} \mathbf{J}(\mathbf{x}) \times \mathbf{B}(\mathbf{x}) \Diff3 \mathbf{x}, \\
	F_i &\approx \int_{V} \epsilon_{ijk} J_j(\mathbf{x}) (B_k(\mathbf{0}) + x_l \partial_l B_k(\mathbf{0})) \Diff3 \mathbf{x} \\
	    &= \epsilon_{ijk} B_k(\mathbf{0}) \int_{V} J_j \Diff3 \mathbf{x} + \epsilon_{ijk} \partial_l B_k(\mathbf{0}) \int_{V} x_l J_j \Diff3 \mathbf{x} \\
	    &= 0 + \epsilon_{ijk} \partial_l B_k(\mathbf{0}) \epsilon_{ljn} m_n \\
	    &= \partial_i B_k(\mathbf{0}) m_k - \partial_k B_k(\mathbf{0}) m_i \\
	    &= \partial_i (m_k B_k) (\mathbf{0}),
\end{align*}
since $\nabla \times \mathbf{B} = 0$. In general, $\mathbf{F} \approx \nabla (\mathbf{m} \cdot \mathbf{B})$. This can also be written as $\mathbf{F} = - \nabla U$, where $U = - \mathbf{m} \cdot \mathbf{B}$ is the potential energy of a magnetic dipole in an external field.

As in the electric case, this in minimized when $\mathbf{m}$ is aligned with $\mathbf{B}$.

\newpage

\section{Electrodynamics}
\label{sec:electrodynamics}

\subsection{Faraday's Law of Induction}
\label{sub:faradays_law_of_induction}

Maxwell's third equation
\[
	\nabla \times \mathbf{E} = - \frac{\partial \mathbf{B}}{\partial t} \tag{M3}\label{eq:m3}
\]
implies that a time-dependent magnetic field must be accompanied by an electric field. This can induce a current to flow in a conductor - a process known as \emph{electromagnetic induction}\index{electromagnetic induction}.

Consider a closed curve $C$ that is the boundary of a time-independent open surface $S$. Integrating (\ref{eq:m3}) over $S$ and using Stokes' theorem,
\[
\int_{C} \mathbf{E} \cdot \diff \mathbf{x} = - \int_{S} \frac{\partial B}{\partial t} \cdot \diff \mathbf{S} = - \frac{\diff}{\diff t} \int_{S} \mathbf{B} \cdot \diff \mathbf{S}
.\]
This is \emph{Faraday's law of induction}\index{Faraday's law of induction} for a static current:
\[
\mathcal{E} = - \frac{\diff \mathcal{F}}{\diff t}
,\]
where
\[
\mathcal{E} = \int_{C} \mathbf{E} \cdot \diff \mathbf{x}
\]
is the \emph{electromotive force}\index{electromotive force} (or emf) around $C$, and
\[
\mathcal{F} = \int_{S} \mathbf{B} \cdot \diff \mathbf{S}
\]
is the \emph{magnetic flux}\index{magnetic flux} through $S$.

Since $\nabla \cdot \mathbf{B} = 0$, the flux $\mathcal{F}$ is the same for any $S$ such that $\partial S = C$, so it can be regarded as the magnetic flux through $C$.

Using $\mathbf{B} = \nabla \times \mathbf{A}$ and Stokes' theorem, we can write the magnetic flux as
\[
\mathcal{F} = \int_{C} \mathbf{A} \cdot \diff \mathbf{x}
,\]
which is invariant under a gauge transformation
\[
\mathbf{A}' = \mathbf{A} + \nabla \chi
.\]
The electromotive force is not actually a force; it is the line integral of the Lorentz force on a particle of unit charge confined to $C$:
\[
\mathcal{E} = \frac{1}{q} \int_{C} \mathbf{F} \cdot \diff \mathbf{x} = \int_{C}(\mathbf{E} + \mathbf{\dot x} \times \mathbf{B}) \cdot \diff \mathbf{x} = \int_{C} \mathbf{E} \cdot \diff \mathbf{x}
,\]
since $\mathbf{\dot x}$ is tangent to $C$ for a particle confined to a time-independent curve $C$.

We will see later that if $C$ coincides with a thin wire of resistance $R$, then the current induced in the wire is $I = \mathcal{E}/R$.

There are several ways in which the magnetic flux through $C$ could change in time:
\begin{itemize}
	\item a magnet is moved near $C$.
	\item a current-carrying circuit is moved near $C$.
	\item the current in a nearby circuit is changed.
\end{itemize}

All these will induce an electromotive force around $C$ and cause a current to flow.

Moreover, we can also generalize Faraday's law for a moving circuit. Let $C(t)$ be a time-dependent closed curve that is the boundary of an open surface $S(t)$. We want to look at how the magnetic flux through $S$,
\[
\mathcal{F} = \int_{S} \mathcal{B} \cdot \diff \mathbf{S}
\]
changes through time. We have
\begin{align*}
	\mathcal{F}(t + \delta t) - \mathcal{F}(t) &= \int_{S(t+\delta t)}\mathbf{B}(\mathbf{x}, t + \delta t) \cdot \diff \mathbf{S} - \int_{S(t)} \mathcal{B}(\mathbf{x}, t) \cdot \diff \mathbf{S} \\
						   &= \int_{S(t+\delta t)} \biggl( \mathbf{B}(\mathbf{x}, t) + \frac{\partial \mathbf{B}}{\partial t} \delta t + \mathcal{O}(\delta t^2) \biggr) \cdot \diff \mathbf{S} - \int_{S(t)} \mathbf{B}(\mathbf{x}, t) \cdot \diff \mathbf{S} \\
						   &= \int_{S(t+\delta t) - S(t)} \mathbf{B}(\mathbf{x}, t) \cdot \diff \mathbf{S} + \int_{S(t)} \frac{\partial \mathbf{B}}{\partial t} \cdot \diff \mathbf{S}\, \delta t + \mathcal{O}(\delta t^2).
\end{align*}
Let $\delta V$ be the volume swept out by $S(t)$ in the time interval $\delta t$. Its boundary is the closed surface $S(t+\delta t) - S(t) + \Sigma$, where $\Sigma$ is the surface swept out by $C(t)$ in time $\delta t$.

\begin{figure}[h]
	\centering
	\caption{Change in Magnetic Flux}
	\label{fig:change_in_mf}
	\begin{tikzpicture}
		\draw (0,0) circle (3 and 1);
		\draw[rotate around={10:(0,2)}] (0,2) circle(3.2 and 0.7);
		\draw (-3,0) -- (-3.154,1.46);
		\draw (3,0) -- (3.154, 2.52);
		\node at (0,0) {\small $S(t)$};
		\node at (0,2) {\small $S(t+\delta t)$};
		\node at (0,-1.3) {\small $C(t)$};
		\node at (2, 1.3) {\small $C(t+\delta t)$};
		\node at (3.3, 1.5) {\small $\Sigma$};
	\end{tikzpicture}
\end{figure}

By (\ref{eq:m2}) and the divergence theorem,
\[
0=  \int_{\partial V} (\nabla \cdot \mathbf{B}) \diff V = \int_{S(t+\delta t) - S(t)} \mathbf{B} \cdot \diff \mathbf{S} + \int_{\Sigma} \mathbf{B} \cdot \diff \mathbf{S}
.\]
To evaluate the last term, parametrize $C$ as $\mathbf{x}=  \mathbf{x}(\lambda, t)$, where $\lambda$ is a parameter around $C$ an element of $C$ is
\[
\diff \mathbf{x} = \frac{\partial \mathbf{x}}{\partial \lambda}\diff \lambda
,\]
and has velocity
\[
\mathbf{v} = \frac{\partial \mathbf{x}}{\partial t}
.\]
In time $\delta t$, it sweeps out the vector area element
\[
\diff \mathbf{S} = \diff \mathbf{x} \times (\mathbf{v} \delta t)
.\]
Thus, we get
\[
\int_{\Sigma}\mathbf{B} \cdot \diff \mathbf{S} = \int_{C} \mathbf{B} \cdot (\diff \mathbf{x} \times \mathbf{v}) \delta t + \mathcal{O}(\delta t^2) = \int_{C}(\mathbf{v} \times \mathbf{B})\cdot \diff \mathbf{x}\, \delta t + \mathcal{O}(\delta t^2)
.\]
Hence we get
\[
\mathcal{F}(t+\delta t) - \mathcal{F}(t) = -\int_{C}(\mathbf{v} \times \mathbf{B}) \cdot \diff \mathbf{x} \, \delta t + \int_{S} \frac{\partial \mathbf{B}}{\partial t} \cdot \diff \mathbf{S} \, \delta t + \mathcal{O}(\delta t^2)
.\]
This gives the first derivative
\begin{align*}
	\frac{\diff \mathcal{F}}{\diff t} &= - \int_{C}(\mathbf{v} \times \mathbf{B})\cdot \diff \mathbf{x} + \int_{S} \frac{\partial \mathbf{B}}{\partial t} \cdot \diff \mathbf{S} \\
					  &= - \int_{C}(\mathbf{v} \times \mathbf{B}) \cdot \diff \mathbf{x} - \int_{S}(\nabla \times \mathbf{E}) \cdot \diff \mathbf{S} \\
					  &= - \int_{C}(\mathbf{E} + \mathbf{v} \times \mathbf{B}) \cdot \diff \mathbf{x}.
\end{align*}
We recover Faraday's law
\[
	\mathcal{E} = - \frac{\diff \mathcal{F}}{\diff t}
,\]
with the redefined electromotive force
\[
\mathcal{E} = \int_{C}(\mathbf{E} + \mathbf{v} \times \mathbf{B}) \cdot \diff \mathbf{x}
.\]
This $\mathcal{E}$ is again the line integral around $C$ of the Lorentz force on a particle of unit charge confined to $C$ (for which the perpendicular components of $\mathbf{\dot x}$ must agree with those of the curve velocity $\mathbf{v}$).

\subsubsection{Lenz's Law}
\label{subsub:lenzs_law}

\emph{Lenz's law}\index{Lenz's law} says that the direction of the induced current is always such as to produce a magnetic field that opposes the change in flux that cause the emf.

\begin{exbox}
	Consider a circular wire in the $xy$-plane. If $B_z$ inside the loop increases in time, then $\mathcal{E} = - \frac{\diff \mathcal{F}}{\diff t} < 0$. This induces a clockwise current ($I < 0$), that generates a magnetic field with $B_z < 0$ inside the loop.

	Hence the minus sign in Faraday's law. This avoids an unstable situation in which the flux grows indefinitely.
\end{exbox}

\begin{figure}[h]
	\centering
	\caption{Lenz's Law}
	\label{fig:lenzs_law}
	\begin{tikzpicture}
		\draw (0,0) circle (3 and 1);
		\draw[->] (-2,-1.4) -- (-2, 1.5);
		\draw[->] (0,-1.5) -- (0,1.7);
		\draw[->] (2,-1.4) -- (2,1.5);
		\node at (0,2) {\small $B_z$};
		\draw[->] (-3,0) arc (180:300:3 and 1);
		\node[below] at (300:3 and 1) {\small $C$};
	\end{tikzpicture}
\end{figure}

\subsubsection{Inductance and Magnetic Energy}
\label{subsub:inductance_and_magnetic_energy}

If a current $I$ around a circuit $C$ generates a magnetic field with flux $\mathcal{F}$, then the \emph{inductance}\index{inductance} of the circuit is defined by
\[
L = \frac{\mathcal{F}}{I}
,\]
and depends only on the geometry of the circuit.

\begin{exbox}
	Consider an ideal solenoid with cross-sectional area $A$ and $N$ turns per unit length. The uniform field $B = \mu_0 N I$ inside the solenoid produces a flux $BA$ per turn, so the inductance per unit length of the solenoid is $\mu_0 N^2 A$.
\end{exbox}

It can be shown that the magnetic flux through a thin wire $C_i$ due to a current $I_j$ around another thin wire $C_j$ is $\mathscr{F}_{ij} = L_{ij} I_j$, where the \emph{mutual inductance}\index{mutual inductance} is
\[
	L_{ij} = \frac{\mu_0}{4 \pi} \int_{C_i} \int_{C_j} \frac{\diff \mathbf{x}_i \cdot \diff \mathbf{x}_j}{|\mathbf{x}_i - \mathbf{x}_j|} = L_{ji}
.\]
When the current $I$ around a circuit $C$ is varied, an emf
\[
\mathcal{E} = - \frac{\diff \mathcal{F}}{\diff t} = - L \frac{\diff I}{\diff t}
\]
is induced. In a small time interval $\delta t$, a charge $\delta Q = I \delta t$ flows around $C$ and the work done on it by the Lorentz force is
\[
\delta W = \mathcal{E} \delta Q = - LI \frac{\diff I}{\diff t} \delta t
.\]
So the rate at which work is done by the current on the electromagnetic field is
\[
- \frac{\diff W}{\diff t} = LI \frac{\diff I}{\diff t} = \frac{\diff}{\diff t} \biggl( \frac{1}{2} LI^2 \biggr)
.\]
Consider reaching a magnetostatic state by building up the current from $0$ to $I$. The energy stored is
\begin{align*}
	U &= \frac{1}{2} LI^2 = \frac{1}{2} I \mathcal{F} = \frac{1}{2} I \int_{C} \mathbf{A} \cdot \diff \mathbf{x} \\
	  &= \frac{1}{2} \int \mathbf{J} \cdot \mathbf{A} \diff V,
\end{align*}
analogous to
\[
U = \frac{1}{2} \int \rho \Phi \diff V
\]
that appears in electrostatics.

Now, using (\ref{eq:m4'}), we have
\[
U = \frac{1}{2 \mu_0} \int (\nabla \times \mathbf{B}) \cdot \mathbf{A} \diff V
,\]
and since $(\nabla \times \mathbf{B}) \cdot \mathbf{A} = \nabla \cdot (\mathbf{B} \times \mathbf{A}) - \mathbf{B} \cdot (\nabla \times \mathbf{A})$, if we take the integral over all space, then the first term gives zero by the divergence theorem, as
\[
|\mathbf{B}| = \mathcal{O} \biggl( \frac{1}{|\mathbf{x}|^3} \biggr), \qquad |\mathbf{A}| = \mathcal{O} \biggl( \frac{1}{|\mathbf{x}|^2} \biggr)
,\]
as $|\mathbf{x}| \to \infty$ for a finite current distribution, leaving
\[
U = \int \frac{|\mathbf{B}|^2}{2 \mu_0} \diff V
\]
as the energy stored in the magnetic field.

\subsection{Ohm's Law}
\label{sub:ohms_law}

In a stationary conductor,
\[
\mathbf{J} = \sigma \mathbf{E}
,\]
where $\sigma$ is the \emph{electrical conductivity}\index{electrical conductivity}. This is not a fundamental physical law, but a constitutive relation, a macroscopic property of a material.

The inverse relation gives
\[
\mathbf{E} = \sigma^{-1} \mathbf{J}
,\]
where $\sigma^{-1}$ is the \emph{resistivity}\index{resistivity}. It is usually denoted as $\rho$, but both $\sigma$ and $\rho$ conflict with notation for charge densities.

A \emph{perfect conductor}\index{perfect conductor} corresponds to the limit $\sigma \to \infty$, so ($\mathbf{E} = 0$), and a \emph{perfect insulator}\index{perfect insulator}  to $\sigma \to 0$ (so $\mathbf{J} = 0$).

\begin{exbox}
	Consider a straight wire of length $L$ in the direction of the unit vector $\mathbf{n}$, and with uniform cross-sectional area $A$ and conductivity $\sigma$. If the electric field is $\mathbf{E} = E \mathbf{n}$, where $E$ is constant, then $\mathbf{J} = \sigma E \mathbf{n}$, and the total current is $I = \sigma E A$.

	The potential difference (voltage) along the wire is
	\[
	V = \int \mathbf{E} \cdot \diff \mathbf{x} = EL = \frac{IL}{\sigma A} = IR
	,\]
	where $R=  \frac{L}{\sigma A}$ is the resistance of the wire.

	Accompanying the resistance of a wire is \emph{Joule heating}\index{Joule heating} (or \emph{Ohmic heating}\index{Ohmic heating}), conversion of electromagnetic energy into heat at the rate $I^2R$.

	If the voltage $V$ is maintained by a battery, then $VI = I^2R$ is the rate at which the emf of the battery ($\mathcal{E} = V$) does work to maintain the current $I$.
\end{exbox}

\begin{figure}[h]
	\centering
	\caption{Ohm's Law in a Wire}
	\label{fig:ohm_wire}
	\begin{tikzpicture}
		\draw (-1, -3) -- (-1, 3);
		\draw (1, -3) -- (1, 3);
		\draw (0,3) circle (1 and 0.3);
		\draw (-1,-3) arc (180:360:1 and 0.3);
		\node at (0,3) {\small $A$};
		\draw[<->] (1.5,-3) -- (1.5,3);
		\node[right] at (1.5, 0) {\small $L$};
		\node at (0.3, -1) {\small $\sigma$};
		\draw[->] (-1.2, -1) -- (-1.2,0);
		\node[above] at (-1.2,0) {$\mathbf{n}$};
	\end{tikzpicture}
\end{figure}

\subsection{Time-dependent Electric Fields}
\label{sub:time_dependent_electric_fields}

Due to time dependence, in electrodynamics we can no longer write $\mathbf{E} = - \nabla \Phi$. But (\ref{eq:m2}) still allows us to write
\[
\mathbf{B} = \nabla \times \mathbf{A}
,\]
and using (\ref{eq:m3}) then gives
\[
\nabla \times \biggl( \mathbf{E} + \frac{\partial \mathbf{A}}{\partial t} \biggr) = 0
.\]
This allows us to write
\[
\mathbf{E} = - \nabla \Phi - \frac{\partial A}{\partial t}
,\]
generalizing the electrostatic expression.

Under a time-dependent gauge transformation
\[
\mathbf{A}' = \mathbf{A} + \nabla \chi, \qquad \Phi' = \Phi - \frac{\partial \chi}{\partial t}
,\]
where $\chi(\mathbf{x}, t)$ is any scalar field, then both $\mathbf{E}$ and $\mathbf{B}$ are unchanged.

\subsubsection{The Displacement Current}
\label{subsub:the_displacement_current}

In magnetostatics we used Ampere's law
\[
\int_{C} \mathbf{B} \cdot \diff \mathbf{x} = \mu_0 \int_{S} \mathbf{J} \cdot \diff \mathbf{S} = \mu_0 I
,\]
or its differential form (\ref{eq:m4'})
\[
\nabla \times \mathbf{B} = \mu_0 \mathbf{J}
.\]
For time-dependent situation, Maxwell's fourth equation,
\[
	\nabla \times \mathbf{B} &= \mu_0 \biggl( \mathbf{J} + \epsilon_0 \frac{\partial \mathbf{E}}{\partial t} \biggr) \tag{M4}\label{eq:m4}
,\]
contains an extra term, the \emph{displacement current}\index{displacement current}.

This is needed, otherwise we would have $\nabla \times \mathbf{J} = 0$, which describes charge conservation in a situation where $\rho$ is constrained to remain constant.

But suppose we place free particles of positive charge in some localized region. Repulsive coulomb forces cause the particles to separate, implying $\nabla \times \mathbf{J} > 0$.

We have seen that the correct form for charge conservation is
\[
	\frac{\partial \rho}{\partial t} + \nabla \cdot \mathbf{J} = 0
.\]
This follows from Maxwell's equations, including the displacement current.

\newpage

\printindex

\end{document}
